Gap formation and stability in non-isothermal protoplanetary discs

### 2014-08-22 paper/model.tex
Paper update

\subsection{Disc model and initial condition}
The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
40\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
surface density profile

@@

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r i}=3\nu/r$ 
with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi i}\simeq r\Omega_k$ 
and $\hat{\nu}\lesssim O(10^{-5})$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.

@@

\subsubsection{Fourier modes}

\subsubsection{Rossby number}

>>>>

\subsection{Disc model and initial condition}
The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
25\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
surface density profile

@@

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r i}=3\nu/r$ 
with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi i}\simeq r\Omega_k$ 
and $\hat{\nu}= 10^{-9}$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.  With this choice of physical viscosity $\hat{\nu}$ the only energy source is through shock-heating (via artificial viscosity) and the $\mathcal{C}$ function when $e\Sigma<e_i\Sigma_i$

@@

\subsubsection{Fourier modes}
Convential Rossby instabilities are characterized by exponetially growing pertibations originating from an axisymetric original state. Though this paper will not deal with formal linear instability scenarios, linear mode measurements will be useful to characterize azimuthal vortex numbers and overdensity values.

The fourier transform of the density field is defined as
\begin{align}\label{fouriertransform}
  \Sigma_m(r,t) = \int_{0}^{2\pi} \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi+\sigma_m \mathrm{t}} \, \mathrm{d}\phi
\end{align}
Here $m$ is the aziumuthal wave frequency and $\sigma_m$ is the angular frequency of the $m \mathrm{th}$ mode which is complex value. Since $\sigma_m$ is a value dependent on $r$, a more useful diagnostic is the mode growth rate defined as 
\begin{align}\label{growth}
  \frac{\partial \langle|\Sigma_m|\rangle_r }{\partial \mathrm{t}}= \gamma \langle|\Sigma_m|\rangle_r 
\end{align}
where $\gamma$ is the real part of the $\sigma_m$ and $\langle|\cdot|\rangle_r$ denotes the radial averaging of the absolute value over the region of interest. By using Eq.~\ref{growth} the growth rates of the linear unstable modes can be found from successive spatial Fourier transforms over an appropriate period of time.

\subsubsection{Rossby number}
The Rossby Number defined as
\begin{align}
  Ro = \frac{{\bf \hat{z}} \cdot \nabla \times \bv - \langle {\bf\hat{z}} \cdot \nabla \times \bv \rangle_{\phi}}{2\langle\Omega\rangle_{\phi}}, 
\end{align}
is a dimensionless number which measures the ratio of interial to Corriolis forces. Here $\langle \cdot \rangle_{\phi}$ denotes azimuthal the averaged quantity. Values of $Ro<0$ correspond to anti-cyclonic rotation with respect to the background motion and thus can be used to quantify area of a vortex and its intensity.

### 2014-08-22 paper/paper.tex
paper update
Paper update

@@

\maketitle
\begin{abstract}
%Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. 
Several observations of transition discs show lopsided
dust-distributions, which is suggestive of a dust-trapping mechanism at work. A
popular explanation is the formation of a large-scale vortex at the
edge of a gap opened by a giant planet. While theoretical models
support this picture, numerical simluations have thus far
employed locally isothermal discs. On the other hand, the theory of the
vortex-forming instability was originally developed for adiabatic
discs. We generalise the study of planetary gap stability to non-isothermal
discs using customised numerical simulations of disc-planet
systems. We include in the energy equation a cooling term with cooling
timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
the Keplerian frequency, and examine the effect of $\beta$ on the
stability of gap edges.

>>>>

\usepackage{array,booktabs,tabularx}
\newcolumntype{R}{>{\centering\arraybackslash}X} % right justified tabularx columns

@@

\maketitle
\begin{abstract}
%Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. 
Several observations of transition discs show lopsided
dust-distributions. This is suggestive of a dust-trapping mechanism at work. A
popular explanation is the formation of a large-scale vortex at the
edge of a gap opened by a giant planet. While theoretical models
support this picture, numerical simluations have thus far
employed locally isothermal discs. On the other hand, the theory of the
vortex-forming instability was originally developed for adiabatic
discs. We generalise the study of planetary gap stability to non-isothermal
discs using customised numerical simulations of disc-planet
systems. We include in the energy equation a cooling term with cooling
timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
the Keplerian frequency, and examine the effect of $\beta$ on the
stability of gap edges. For measurments reminsent of linear stability, gap edges stabilize with increasing $\beta$ as dominate modes and growth rates decrease in value. In the non-linear limit of planet-vortex interactions, large scale overdense vortices were found to last for 1000's of orbits in a quasi-steady state characterized by sudden disspation on the order of 100's or planet orbits. Vortex lifetimes were found to have a non-monoticly dependence on $\beta$ and an optimum value for cooling rate was found to maximize the lifetime.

### 2014-08-22 paper/results.tex
Paper update

%The goal of these simulations is to compliment the disk-planet
%simluations presented later. 
These `planet-off' simulations are not linear stability calculations
because our cooling term restores the initial temperature profile
corresponding to $H/r=0.05$, rather than the heated gap edge.

@@

\subsection{Gap structure}
{\bf plots of the gap structure before instability sets in, e.g. at
  t=30. surface density perturbation, gen. vortensity perturbation,
  aspect-ratio. compare diff. beta. gaps become more shallow with
  increasing cooling time (but not significantly - not enough time for
  full gap formation). generalized vortensity: extrema smoothed out
  with increasing cooling - expect more stability. aspect-ratio:
  significant heating. increase in H so expect larger vortices
}

\subsection{Linear modes and growth rates}
{\bf present `planet-off' simulations. one `fourier mode v.s. time'
  plot to show growth of linear instability. do an adiabatic case (or
  extremely long cooling time) to see the effect of heated gap edge
  (i.e. temp doesn't go back to t=0 value too quickly)
  compare growth rate and dominant
  m as function of beta (table). 2D figs to contrast (also used to
  show it's the minimum in generalized pv that goes unstable.    
  result: increasing cooling time makes the gap
  more stable, and favors lower m. note: the `basic state' should be
  the system at t=30 after azimuthal average. linear results should
  have small perturbations. 
}

\subsubsection{Nearly adiabatic disks}
{\bf a case with extrmely long cooling time (or adiabatic run) in
  order to look at effect of heated gap edge.  prelim result: not
  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
  purely adiabatic disk. change in $h$ not important for linear
  perturbations (but important for setting up the basic state). 
}
%% We first describe results for weakly self-gravitating
%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
%% vortex instablity \citep{valborro07}. These parameter values
%% are typical for disc-planet simulations and yield stable/steady 
%% gap profiles. This will be useful reference cases to  
%% understand the unstable cases considered later.

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time} 
{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
  to plot vortex lifetime v.s. beta? table: 
  averaged quantities over quasi-steady state: aspect-ratio (to
  compare with fu at al), rossby number (vortex strength
  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
  only inviscid cases. describe evolution of one case. main
  conclusion: longer vortex lifetime with increasing cooling time (up
  to some optimal timescale). vortex death: induced-shock and/or
  smoothing the gap edge. describe simulation setup, resolution?
  should mention that results consistent with lower-resolution prelim
  runs. maybe torques? 
}

Long term simulation of vortices were done for $\tilde{\beta}=0.1,1.0$ and
 $\hat{\nu}=10^{-6},10^{-8},10^{-9}$ up to a total of 2000 orbits. Values
 of higher $\tilde{\beta}$ resulted generally in longer vortex lifetimes.
 The inviscid cases also were characterized by a sudden death of vortices
 while the higher viscosity case showed a gradual decrease in mode amplitudes
 attributed to the vortices.

Short term simulations of the linear phase of instability growth also distinguished
 the different effects of cooling rates. Simulations with higher
 $\tilde{\beta}$ were characterized by slower growth rates centered at the
 $m=2$ mode, lower density pertibations and vortex locations, and a more unitary aspect
 ratio for the vortices.

 \begin{figure}
   \includegraphics[scale=.42]{figures/stability_vis9betalow.ps}
   \caption{Mode intensity relative to $m=0$ for $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$.}
 \label{stability_vis9lowb}
 \end{figure}

\begin{figure}
   \includegraphics[scale=.42]{figures/stability_vis9betamed.ps}
   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=1.0$. }
 \label{stability_vis9medb)}
 \end{figure}

@@

\begin{figure}
   \includegraphics[scale=.60]{figures/analysis_sigma50lowb.ps}
   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$. }
 \label{shortterm_lowb)}
 \end{figure}

\begin{figure}
   \includegraphics[scale=.60]{figures/analysis_sigma50highb.ps}
   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=10.0$. }
 \label{shortterm_highb)}
 \end{figure}

>>>>

Simulations were resolved with grid divisions of $(N_r,N_{\phi})=(1024,2048)$ for time up to $200P_0$ allowing high resolution and detection of high $m$ structures. Cases of $\tilde{\beta}=0.1,1.0,10.0$ were analyzed to provide a range of fast, moderate, and slowly cooled disks.

%The goal of these simulations is to compliment the disk-planet
%simluations presented later. 
These `planet-off' simulations are not linear stability calculations
since the cooling term in our energy equation restores the initial temperature profile corresponding to $H/r=0.05$, rather than the heated gap edge. Linear stability calculations and adiabatic cases will be further discussed in section~\ref{adiabatic_section}.

@@

\subsection{Gap structure}
%{\bf plots of the gap structure before instability sets in, e.g. at
%  t=30. surface density perturbation, gen. vortensity perturbation,
%  aspect-ratio. compare diff. beta. gaps become more shallow with
%  increasing cooling time (but not significantly - not enough time for
%  full gap formation). generalized vortensity: extrema smoothed out
%  with increasing cooling - expect more stability. aspect-ratio:
%  significant heating. increase in H so expect larger vortices
%}

Since stability of the gap is of interest we analyze the intial state formed by the planet-disk intereactions. The intial gap profiles after the averaging procedure are shown in Fig \ref{intial1D}. Quicker cooling rates (lower $\tilde{\beta}$) result in deeper density gradients and more shallow gaps in the disk but never significant enough to form a clear gap in this time period. Generalized vortensity profiles show smoothening with larger cooling rates with the extrema becoming less pronounced. Charactereristicly for such disk-planet Rossby vortices the generalized vortensity minima correspond to density maxima in the disk. Larger $\tilde\beta$ values also as expeced result in higher disk aspect ratios $H/r$, which is a value directly correlated to temperature in our model.

 For a planet to carve a gap in the surrounding material it has been shown that $q>H^3$ is needed \citep{crida06}. This indicates that for hotter disks, which have respectively higher $H$, it becomes more difficult by a cubic dependence for planet induced gaps to be formed. This is seen in our simulations as changes in $\tilde{\beta}$ result in changes of $h$ by less than $20\%$ yet gap depth almost doubles. Since generalized vortensity extrema are also correspondingly less pronounced and density gradients become less large we expect that the stability of gaps to increase with larger $\tilde{\beta}$.

\begin{figure}
  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
    2cm 0cm 0cm]{figures/compare_sigma}
  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
    2cm 0cm 1cm]{figures/compare_gvortensity}
\includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
    0.5cm 0cm 1cm]{figures/compare_aspectratio}
  \caption{Gap profiles for the intial partial gap opened before instability emerges at $t=30P_0$ for quick (solid line), moderate (dashed line), and slow cooling (broken line). Relative density pertibation (top), generalized vortensity pertibation (middle) and disk aspect ratio (bottom) are shown. \label{intial1D}}
\end{figure}

%\begin{tabularx}{0.4\textwidth}{l*{10}{R}} \toprule
%  \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
%  m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
 % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
%\end{tabularx}

%\begin{tabularx}{0.4\textwidth}{l*{5}{R}} \toprule
%  \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
%  m                    & 1 & 2 & 3 & 4 & 5  \\ 
%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
%\end{tabularx}

\begin{table}
  \centering
  \caption{Dominant mode and growth rates for $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling) values during `planet-off' simulations \label{modetable}}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
      \toprule
      m & $\gamma10^2/\Omega(r_p)$ \\
      \midrule
      6 & 7.3 \\
      7 & 7.8 \\
      8 & 7.9 \\
      9 & 7.9 \\
      10 & 6.8 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
      \toprule
      m & $\gamma10^2/\Omega(r_p)$ \\
      \midrule
      3 & 2.0 \\
      4 & 2.2 \\
      5 & 2.3 \\
      6 & 1.6 \\
      7 & 1.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
      \toprule
      m & $\gamma10^2/\Omega(r_p)$ \\
      \midrule
      1 & 1.1 \\
      2 & 1.6 \\
      3 & 1.7 \\
      4 & 1.2 \\
      5 & 0.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
\end{table}

\subsection{Linear modes and growth rates}\label{linear}
%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
%  plot to show growth of linear instability. do an adiabatic case (or
%  extremely long cooling time) to see the effect of heated gap edge
%  (i.e. temp doesn't go back to t=0 value too quickly)
%  compare growth rate and dominant
%  m as function of beta (table). 2D figs to contrast (also used to
%  show it's the minimum in generalized pv that goes unstable.    
%  result: increasing cooling time makes the gap
%  more stable, and favors lower m. note: the `basic state' should be
%  the system at t=30 after azimuthal average. linear results should
%  have small perturbations. 
%}
After the azimuthal averaging, the small pertibation excites rapidly growing instabilities by the RWI. We analyticly charcterize these modes with quantities $m$ and $\gamma$ as defined by Eq.~\ref{fouriertransform} and Eq.~\ref{growth}. Table \ref{modetable} contains the results of the mode growth rates for a variety of cooling rates during this semi-linear phase. Results indicate that as $\tilde{\beta}$ was increased from $ 0.1\rightarrow10.0$ the dominant azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the respective growth rate for mode decreased from $ \gamma/\Omega(r_p)=0.079 \rightarrow 0.017$. Snapshots of the instability in $r-\phi$ plane for the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.

These `planet-off' simulations, which are reminiscient of linear stability measurements, show that stability of gap edges increase with longer cooling rates (ie hotter gaps). This is as expected from the intial gap profiles and the gap forming criteria discussed in the previous section.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
    0cm 0cm 0cm]{figures/linear_stability}
  \caption{Log plot of the azimuthal Fourier mode of disk surface density non-dimenionlized by the background mode, during the `planet-off' simulations for $\tilde{\beta}=10$. Colors correspond to different $m$ values. The m=3 component of the fourier transform clearly is most dominant during this growth phase of the instability and has corresponding $\gamma=0.017\Omega(r_p)$. \label{linearmodes}}
\end{figure}

\begin{figure*}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_lowb}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_medb}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_highb}
  }  
  \caption{$r-\phi$ plots of the generalized vortensity pertibation for cases of $\tilde{\beta}=0.1,1.0,10.0$ (left,middle,right) during the linear phases of growth. The number of visible vortices decrease as $\tilde{\beta}$ increases and snapshot times correspondingly increase to capture the appropriate phase of the growth. \label{2Dlinear}}
\end{figure*}

\subsubsection{Nearly adiabatic disks}
\label{adiabatic_section}
%{\bf a case with extrmely long cooling time (or adiabatic run) in
%  order to look at effect of heated gap edge.  prelim result: not
%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
%  purely adiabatic disk. change in $h$ not important for linear
%  perturbations (but important for setting up the basic state). 
%}
During the `planet-off' simulations the analysis is not truly a linear stability measurement as we do not capture the effects of a heated disk on the instability growth. Since $t_{\mathrm{cool}}<t_{\mathrm{instability}}$ the disk manages to cool back to its intial value of $h=0.05$ before the instability emerges and thus we do not have a steady background state to form a linear stability problem. To compare our 'planet-off' analysis with a true linear stability analysis we ran a simulation  with $\tilde{\beta}=100.0$ which corresponds with an almost adiabtaic case. In this simulation the cooling rate is slow enough that the gap temperature profile changes only marginally over the timescale of the growth of the instability. The results show that the linear stability of the $\tilde{\beta}=100.0$ cases almost exactly equivilant to the case of $\tilde{\beta}=10.0$ in terms of mode growth rates and gap profiles. Since the disk heats up only to values $h\simeq0.06$ in the adiabatic case, the difference in growth rates between this and the original disk temperature of $h=0.05$ changes almost insignificantly as shown by linear theory \citep{li00}. This indicate that a continuously hot gap edge is not as important for development of vortices as the heating effects on the formation of the intail gap state.

%% We first describe results for weakly self-gravitating
%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
%% vortex instablity \citep{valborro07}. These parameter values
%% are typical for disc-planet simulations and yield stable/steady 
%% gap profiles. This will be useful reference cases to  
%% understand the unstable cases considered later.

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time} 
%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
%  to plot vortex lifetime v.s. beta? table: 
%  averaged quantities over quasi-steady state: aspect-ratio (to
%  compare with fu at al), rossby number (vortex strength
%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
%  only inviscid cases. describe evolution of one case. main
%  conclusion: longer vortex lifetime with increasing cooling time (up
%  to some optimal timescale). vortex death: induced-shock and/or
%  smoothing the gap edge. describe simulation setup, resolution?
%  should mention that results consistent with lower-resolution prelim
%  runs. maybe torques? 
%}

Long term simulation of edge vortices were done for $\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of $2000P_0$. Planets were left free to interact dynamiclly with the disk and vortices after $t=30P_0$ as apposed to previous section. These simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in order for them to be computationally economical. An open boundry condition was used in the outer edge of the simulation region.

After the semi-linear growth of vortices, vortex merger effects take hold and by 150$P_0$ for all $\tilde\beta$ values, only $m=1$ vortex modes exist. Plots of the amplitude of the $m=1$ mode for the different $\tilde\beta$ cases are shown in Fig~\ref{lifetimeplot}. These instabilities stay in quasi-steady states for longer than a 1000 of orbits in most cases, in which they grow in intensity and accumulate density. The growth of these vortices are supplied by the continuous generation of vorticity by the planet-disk interactions. Eventually the $m=1$ amplitudes are also characterized by sudden large drops of vortex intensity which happen over timescales of 20$P_0$. After these drops the instability is never seen to reform at such large over densities and density bumps are never larger than the planet induced shocks. We designate the time of these amplitude drops as the lifetime of the quasi-steady vortex.

Rossby values have dynamics that are coupled with the lifetime and $m=1$ amplitude of the vortices. As the vortex forms it has a charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic motion. As the $m=1$ amplitude grows the relative spin of the vortex contiously grows along with it, reaching values of $Ro\approx-0.4$. As the vortex dies off as described above the associated Rossby number of the vortex accordingly quickly approaches 0.

The sudden dissipation of these vortices seem to be a result of the vortices shocking the system. Measurements of the Mach number near the vortices show large increases in value when the vortex is dissipating which indicates that the system is more susceptible to shocks. Also gradients of surface density are seen to spike up in wake like features around the vortex during dissipation which is also indicative of shocks. Of note is also that the large overdensity the vortex creates during it's quasi-steady state distorts the background disk. So as the vortex grows in intensity it alters the intial gap structure that formulated its growth.

Contradictory to our `planet-off' simulations, vortex lifetimes did not monotonicly depend on intial edge stability and from our simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes the vortex lifetime as seen in Fig~\ref{lifetimeplot}. Increasing $\tilde\beta$ correspondingly increased the lifetime of the vortex up to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that lasted beyond the simulation time of 2000$P_0$. Cooling rates with $\tilde\beta>5.0$ showed lifetimes that were well below the max lifetime. By analyzing the aspect ratio of the corotation region around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk corresponds to $h\approx0.06$ which is a result similarily found by \citet{fu14} to be the optimum lifetime for long-term simulations of isothermal disks with varying intial temperature profiles $h$.

Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in the the non-montonic form of the lifetime. As the $\tilde\beta$ increases so does the stability of the gap edge as seen in section~\ref{linear}, however as $\tilde\beta$ increases so does the $c_{\mathrm{iso}}$ which is known from theory to effect the growth of instability \citep{li00}. Thus increasing $\tilde\beta$ increases the sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to a certain point in which the effects of the more stable gap edge becomes a significant effect. The correlation of growth rate and sound speed is non-negligible as in previous `planet-off' case as the disk now can reach values of $h\simeq0.08$ for very slow cooling and the importance of the effect can be imagined to accumulate over the now considerably longer timescale. These effects dictate when then the vortex reaches the critical value that causes disspation and hence the lifetime fo the vortex.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$ background mode for longterm non-linear simulations corrsponding to vortex lifetime. The longest living mode correpsonds to $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{2-D snapshot in polar form of relative density pertibation for $\tilde\beta=0.1$ vortex during quasi-steady state at $t=P_0$. Large scale non-axisyemetric overdensity can be seen. \label{Vortex2D}}
\end{figure}

@@
